### 2019-09-10 ch1_introduction/content.tex
update
updaze

\section{The cell}
The cell is the smallest unit of life, of which all known living organisms are composed. Every cell houses a plethora of biomolecular processes that allow it to adapt to changes in their environment continuously. It can be very challenging to comprehend the cellular response to a signal due to the dynamic nature of these processes. A reductionist approach to understanding a complex biological system is to study the biochemical components which it is comprised of \cite{brigandt_reductionismbiology_2017}.

Reductionist biologists are delighted by recent advances in experimental technologies that permit measuring the abundance of thousands of different biochemical molecules in tens of thousands of individual cells.  Observing the biomolecular insides of cells in this manner will ultimately provide fundamental insights into the processes that govern these cells and help uncover novel approaches for diagnosing and treating disease. Every coin has its flip side, however, and in this case, it is that the amount of data generated from these experiments is not analysable by hand.

@@

The sheer volume of the data generated from such highly-integrative and high-throughput experiments are not the only reason why they are so challenging to interpret. Namely, the data inherently also suffers from batch effects arising from differences between donors and labs, and also contains high levels of noise arising from the experimental profiling techniques used \cite{hon_humancellatlas_2018}. Biologists thus turn to computer scientists\footnote{or computational biologists turn to themselves} to develop new tools to tackle these problems and help biologists extract meaningful biological insights from the data.

@@

\subsection{The origins of life and the RNA world}
The discovery of the double helix shape of Deoxyribonucleic Acid (DNA) \cite{watson_molecularstructurenucleic_1953} is often considered the pivot point in our understanding of the origins of life and evolution. A modern high-school student can plausibly be expected to know that DNA serves as a medium for storing the genetic information required to reproduce a whole organism. With other words, the DNA of an organism contains the complete set of instructions required to build all of the biomolecular machinery present in its body. The magnitude of this discovery is reflected in our language and culture alike; with sayings such as "It's in your DNA.", or usage of its shape in countless illustrations or artworks (Figure \ref{fig:fsvm}).

\begin{figure}
	\centering\includegraphics[width=.5\linewidth]{fig/fsvm3} % TODO: change image
	\caption{
		\textbf{A prominent display of the double helix shape at the VIB FSVM building.} % TODO: expand description
	}
	\label{fig:fsvm}
\end{figure}

@@

\begin{figure}
	\centering\includegraphics[width=.5\linewidth]{fig/rnaworld} % TODO: change image
	\caption{
		\textbf{RNA world.} The postulated rise and fall of the RNA world during the evolution of life, from early self-replicating RNA to complex, RNA-controlled metabolism, to the invention of translation, followed by diversification of all modern branches of life. Image from Horning (2011) \cite{horning_rnaworld_2011}. % TODO: expand the description
	}
	\label{fig:rnaworld}
\end{figure}

\subsection{The Central Dogma}
The Central Dogma is a set of processes present which govern the general flow of genetic information in almost all existant living cells. In short, it states that DNA codes for RNA, which in turn codes for proteins. In this work, we assume the main processes involved in the Central Dogma are replication, transcription, splicing and translation (Figure 3).

\textbf{Replication} is the process of duplicating DNA, which allows a cell to divide such that the resulting daughter cells retain complete copies of its genetic information. DNA consists of four different so-called nucleobases named adenine (A), cytosine (C), guanine (G), and thymine (T). Each strand of the double helix structure of DNA is a linear chain of nucleobases. The two strands are held together by hydrogen bonds since adenine can form three hydrogen bonds with thymine and cytosine can form two hydrogen bonds with guanine. Since each nucleobase can only bond with one other nucleobase, one strand is complementary to the other. Thus a strand of DNA can be represented by a sequence of letters, for example, "ACTCGGTTTAGCA". % This could be part of the DNA explanation.

A stretch of DNA that contains a genetic blueprint for a particular molecule is called a gene, and the collection of an organism's genes is called its genome. \textbf{Transcription} is the process of synthesising an RNA molecule from a gene, and the resulting molecule is called a transcript. RNA is similar to DNA but differs in several ways; most notably all thymine nucleobases are replaced with uracil (U), and RNA molecules consist of only one strand. Due to its single-strandedness, RNA is less stable and will break down faster. Single-strandedness also allows some types of RNA (e.g. transfer RNA, ribosomal RNA) to form more complex three-dimensional structures by having certain regions bind to other regions of the strand. This work only considers messenger RNA (mRNA). mRNAs are transcribed from protein-coding genes, meaning that the mRNAs will lead to the production of protein molecules through the process of translation. 
% TODO: mention that there are about 20'000 protein-coding genes in the human body.

\textbf{RNA splicing} is a process that occurs in almost all organisms (but not all) and results in the removal of specific regions in the RNA molecule. Introns are the regions removed by this process, and exons the regions that remain. The main functionality of splicing is to be able to create multiple variants of the same product, which can affect the enzymatic properties or localisation of the resulting product \cite{kelemen_functionalternativesplicing_2013}.

During \textbf{translation}, a chain of amino acid is synthesised from an mRNA transcript. Every three nucleobases are translated into one of 21 different amino acids. The resulting chain of amino acids is folded up into a protein, the structure of which is determined by the sequence of different amino acids in the chain. In turn, its structure determines the functionality of the protein, which includes catalysing biochemical reactions, providing structure, and transportation of molecules.

@@

\textit{Homo sapiens} like to categorise everything they encounter and so too have they conceptualised groups of cells called "cell types" according to their functionality. The concept of cell types eases reasoning about all aspects of biology, for instance, which cell types turn into (differentiate) or communicate with which other cell types, or how a cell type responds to a specific stimulation. Cells can be highly specialised toward performing a particular function (e.g. memory B cells accelerate immune response by remembering previously encountered pathogens), while other cells maintain a strong ability to differentiate into other cell types.

@@

\subsection{Cell dynamics and gene regulation}
If cells are dynamic entities and can gradually produce the molecules needed to acquire new functionality, what is the process by which this happens? The mechanism by which this happens is called gene regulation. Some proteins (or other molecules such as micro RNAs) are capable of determining the rate at which a gene is transcribed (transcription rate). Such proteins are called transcription factors (TFs), and the genes it regulates are called its targets. Typically, one TF will regulate the transcription rate of many targets. % TODO: download database and calculate some statistics.

@@

The earliest approaches for measuring the abundance of particular molecules in \emph{single cells} used the preferred instrument of every stereotypical biologist: the microscope. Since it was developed by Coons et al. in 1941, immunohistochemistry (IHC) has been instrumental in visualising antigen-antibody proteins \cite{coons_immunologicalpropertiesantibody_1941}. In many multicellular organisms, antibodies and antigens serve as crucial communication tools as part of the organism's immune system. A cell can present a particular type of antigen on its cell surface, which allows a particular type of antibody to bind to it.

@@

@@

\begin{figure}
	\centering
	A. \includegraphics[width=.45\linewidth]{fig/profiling_convergence} 
	% also figure with cell numbers
	C. \includegraphics[width=.45\linewidth]{fig/multimodal} 
	% TODO: change image
	\caption{
		A. \textbf{Convergence of "Omics" Biology and Single-Cell Biology.} Technology that allows researchers to obtain genome-wide information from single cells is extending the boundaries of a field that has thus far been limited to the analyses of a select gene in eukaryotes. Image from Junker and van Oudenaarden (2014) \cite{junker_everycellspecial_2014}. 
		B. 
		C. scmultiomics \cite{moudgil_multimodalscrnaseq_2019}. 
		% TODO: rewrite the description
	}
	\label{fig:profiling_convergence}
\end{figure}

@@

\begin{figure}
	\centering\includegraphics[width=.5\linewidth]{fig/singlecell_tools}
	\centering\includegraphics[width=.3\linewidth]{fig/scrnatools}
	% TODO: change image
	\caption{\textbf{A. Single-cell omics allows for many new types of computational approaches.} Figure adapted from Wagner et al. (2016) \cite{wagner_revealingvectorscellular_2016}.
		B. Zappia et al. (2018) \cite{zappia_exploringsinglecellrnaseq_2018}
		% TODO: rewrite the description
	}
	\label{fig:comp_tools}
\end{figure}

@@

\begin{figure}
	\centering\includegraphics[width=\Largefigure]{fig/trajectory_inference_slim} % TODO: Update figure to match text!!!!
	\caption{
		Applications of single-cell trajectory inference methods. (A) Single-cell omics data appropriate for TI can be both obtained from an unsynchronised population of single cells (snapshot data) but also from synchronised cell populations. (B) UPDATE! (C) UPDATE!
	}
	\label{fig:trajectory_inference}
\end{figure}

@@

\begin{figure}
	\centering
	\includegraphics[width=\Largefigure]{fig/ni_bulk_vs_sc}
	\caption{Bulk expression data return the average expressions of genes among large numbers of cells. In order to infer regulatory networks from this type of data, multiple bulk profiles (resulting from time series or perturbation experiments) are required. On the other hand, sequencing the transcriptome at the single-cell level uncovers the high variability among cells, providing the necessary information to infer gene regulatory networks directly.} % TODO: update label
	\label{fig:bulk_vs_singlecell}
\end{figure}

@@

\subsection{Co-author publications}
\begin{itemize}
	\item Decock A, Ongenaert M, \textbf{Cannoodt R}, Verniers K, De Wilde B, Laureys G, Van Roy N, Berbegall AP, Bienertova-Vasku J, Bown N, Clément N. Methyl-CpG-binding domain sequencing reveals a prognostic methylation signature in neuroblastoma. Oncotarget. 2016 Jan 12;7(2):1960.
	\item Van Cauwenbergh C, Van Schil K, \textbf{Cannoodt R}, Bauwens M, Van Laethem T, De Jaegere S, Steyaert W, Sante T, Menten B, Leroy BP, Coppieters F. arrEYE: a customized platform for high-resolution copy number analysis of coding and noncoding regions of known and candidate retinal dystrophy genes and retinal noncoding RNAs. Genetics in Medicine. 2017 Apr;19(4):457.
	\item Claeys S, Denecker G, \textbf{Cannoodt R}, Kumps C, Durinck K, Speleman F, De Preter K. Early and late effects of pharmacological ALK inhibition on the neuroblastoma transcriptome. Oncotarget. 2017 Dec 5;8(63):106820.
	\item Depuydt P, Boeva V, Hocking TD, \textbf{Cannoodt R}, Ambros IM, Ambros PF, Asgharzadeh S, Attiyeh EF, Combaret V, Defferrari R, Fischer M. Genomic amplifications and distal 6q loss: novel markers for poor survival in high-risk neuroblastoma patients. JNCI: Journal of the National Cancer Institute. 2018 Mar 5;110(10):1084-93.
	\item Scott CL, T'Jonck W, \ldots, \textbf{Cannoodt R}, Saelens W \ldots, Guilliams M. The transcription factor ZEB2 is required to maintain the tissue-specific identities of macrophages. Immunity. 2018 Aug 21;49(2):312-25.
	\item Saelens W, \textbf{Cannoodt R}, Saeys Y. A comprehensive evaluation of module detection methods for gene expression data. Nature communications. 2018 Mar 15;9(1):1090.
	\item Todorov H, \textbf{Cannoodt R}, Saelens W, Saeys Y. Network Inference from Single-Cell Transcriptomic Data. InGene Regulatory Networks 2019 (pp. 235-249). Humana Press, New York, NY..
	\item Van den Berge K, De Bezieux HR, Street K, Saelens W, \textbf{Cannoodt R}, Saeys Y, Dudoit S, Clement L. Trajectory-based differential expression analysis for single-cell sequencing data. BioRxiv. 2019 Jan 1:623397.
	\item Weber LM, Saelens W, \textbf{Cannoodt R}, Soneson C, Hapfelmeier A, Gardner PP, Boulesteix AL, Saeys Y, Robinson MD. Essential guidelines for computational method benchmarking. Genome biology. 2019 Dec;20(1):125.
\end{itemize}

@@

%library(dlstats)
%library(tidyverse)
%
%cran <- c("babelwhale", "dyndimred", "dynparam", "dynutils", "GillespieSSA", "GillespieSSA2", "incgraph", "princurve", "proxyC", "qsub", "SCORPIUS", "devtools", "mlr", "ParamHelpers", "remotes", "rlang")
%bioc <- c("ClusterSignificance", "monocle", "slingshot", "splatter")
%
%dl_cran <- dlstats::cran_stats(cran)
%dl_bioc <- dlstats::bioc_stats(bioc)
%
%dls <- bind_rows(dl_cran, dl_bioc %>% select(start, end, package, downloads = Nb_of_downloads)) %>% 
%mutate(role = ifelse(package %in% c("devtools", "mlr", "ParamHelpers", "remotes", "rlang", bioc), "Contributor", "Author"))
%
%summ <- dls %>% filter("2019-08-01" <= start) %>% group_by(package) %>% summarise(downloads = sum(downloads) / 23 * 365)
%
%summ <- dls %>% filter("2018-07-01" <= start, end <= "2019-07-30") %>% group_by(package) %>% summarise(downloads = sum(downloads))

\begin{table}[ht!]
	\caption{\textbf{Contibutions to open-source software.} Following abbreviations denote the role within the package: aut Author, ctb Contributor. Yearly download statistics are based on the number of downloads between 2019-08-01 and 2019-08-23. CRAN download statistics are retrieved from the Rstudio CRAN mirror only; other CRAN mirrors do not track download statistics. For Github repositories, no download statistics could be retrieved. } \label{tab:packages}

@@

		\cranpkg{babelwhale} & aut & CRAN & 3491 & Interacting with Docker and Singularity containers \\

@@

		\cranpkg{dyndimred} & aut & CRAN & 571 & Applying dimensionality reduction methods \\

@@

		\cranpkg{dynparam} & aut & CRAN & 2539 & Creating meta-information for parameters \\

@@

		\cranpkg{dynutils} & aut & CRAN & 4570 & Common functionality for the dynverse packages \\

@@

		\cranpkg{GillespieSSA} & aut & CRAN & 7855 & Gillespie's Stochastic Simulation Algorithm (SSA) \\
		\cranpkg{GillespieSSA2} & aut & CRAN & 2253 & Gillespie's Stochastic Simulation Algorithm for Impatient People \\

@@

		\cranpkg{incgraph} & aut & CRAN & 2809 & Incremental graphlet counting for network optimisation \\
		\cranpkg{princurve} & aut & CRAN & 24'439 & Fits a principal curve in arbitrary dimension \\
		\cranpkg{proxyC} & aut & CRAN & 112'753 & Computes proximity in large sparse matrices \\
		\cranpkg{qsub} & aut & CRAN & 2698 & Running commands remotely on gridengine clusters \\
		\cranpkg{SCORPIUS} & aut & CRAN & 4824 & Inferring developmental chronologies from single-cell RNA sequencing data \\ \hline\hline
		\biocpkgl{ClusterSignificance}{Clus\-ter\-Sig\-nif\-i\-cance} & ctb & Bioc & 571 & Assess if class clusters in dimensionality reduced data representations have a separation different from permuted data \\
		\cranpkg{devtools} & ctb & CRAN & 3'254'816 & Tools to make developing R packages easier \\

@@

		\cranpkg{mlr} & ctb & CRAN & 141'192 & Machine Learning in R \\
		\biocpkg{monocle} & ctb & Bioc & 29'121 & Clustering, differential expression, and trajectory analysis for single-cell RNA-Seq \\
		\cranpkg{ParamHelpers} & ctb & CRAN & 101'137 & Helpers for Parameters in Black-Box Optimization, Tuning and Machine Learning \\

@@

		\cranpkg{remotes} & ctb & CRAN & 3'476'435 & R package installation from remote repositories, including GitHub \\
		\cranpkg{rlang} & ctb & CRAN & 10'314'725 & Functions for base types and core R and tidyverse features \\

@@

		\biocpkg{slingshot} & ctb & Bioc & 9538 & Tools for ordering single-cell sequencing \\
		\biocpkg{splatter} & ctb & Bioc & 2999 & Simple simulation of single-cell RNA sequencing data \\

>>>>

\section{The cell}
The cell is the smallest unit of life, of which all known living organisms are composed. Every cell houses a plethora of biomolecular processes that allow it to adapt to changes in its environment continuously. It can be very challenging to comprehend the cellular response to a signal due to the dynamic nature of these processes. A reductionist approach to understanding a complex biological system is to study the biochemical components which it is comprised of \cite{brigandt_reductionismbiology_2017}.

Recent advances in experimental technologies are playing a crucial role in reductionist biology, allowing to measure the abundance of thousands of different biochemical molecules in tens of thousands of individual cells. Observing the biomolecular insides of cells in this manner can ultimately provide fundamental insights into the processes that govern these cells and help uncover novel approaches for diagnosing and treating disease. Every coin has its flip side, however, and in this case, it is that the amount of data generated from these experiments is not analysable by hand.

@@

%%%% HT: I feel like you're only going to focus on preprocessing, when reading this section. Maybe add a second paragraph mentioning that separating technical and biological noise from actual biological processes in the data is not the only challenge, and that there is also a necessity of new computational tools to start generating more complex models, integrating the information from single cell high sequencing experiments in a much more refined way?
%%% RC: In general, these two paragraphs are not telling the reader a lot about where this thesis is going

The sheer volume of the data generated from such highly-integrative and high-throughput experiments are not the only reason why they are so challenging to interpret. 
Namely, the data contains high levels of noise arising from inherent biomolecular stochasticity in the cells and from the experimental profiling techniques uses, as well as batch effects arising from differences between donors and labs \cite{hon_humancellatlas_2018}. Biologists thus turn to computer scientists\footnote{or computational biologists turn to themselves} to develop new tools to tackle these problems and help biologists extract meaningful biological insights from the data.

@@

\subsection{The origins of life and the RNA world}
The discovery of the double helix shape of Deoxyribonucleic Acid (DNA) \cite{watson_molecularstructurenucleic_1953} is often considered the pivot point in our understanding of the origins of life and evolution. By now, it is well known that DNA serves as a medium for storing the genetic information required to reproduce a whole organism. With other words, the DNA of an organism contains the complete set of instructions required to build all of the biomolecular machinery present in its body.
The magnitude of this discovery is reflected in our language and culture alike; with sayings such as "It's in your DNA.", or usage of its shape in countless illustrations or artworks (Figure \ref{fig:fsvm}).
%% HT: I'm not sure if the magnitude of the DNA discovery is best reflected in sayings and double helix monuments it sounds a bit light to me, compared to the fact that you can identify geneticel targets of a disease and Crispr cas it for instance, but that's a personal opinion, feel free to ignore it :)

\begin{figure}[htb!]
	\centering\includegraphics[width=.5\linewidth]{fig/fsvm3} % TODO: change image
	\caption{
		\textbf{A prominent display of the double helix shape at the VIB FSVM building.} % TODO: expand description
	}
	\label{fig:fsvm}
\end{figure}

@@

\begin{figure}[htb!]
	\centering\includegraphics[width=.5\linewidth]{fig/rnaworld} % TODO: change image
	\caption{
		\textbf{RNA world.} The postulated rise and fall of the RNA world during the evolution of life, from early self-replicating RNA to complex, RNA-controlled metabolism, to the invention of translation, followed by diversification of all modern branches of life. Image from Horning (2011) \cite{horning_rnaworld_2011}. % TODO: expand the description
	}
	\label{fig:rnaworld}
\end{figure}

\subsection{The Central Dogma}
The Central Dogma is a set of processes present which govern the general flow of genetic information in almost all existent living cells. In short, it states that DNA codes for RNA, which in turn codes for proteins.
In this work, we assume the main processes involved in the Central Dogma are replication, transcription, splicing and translation (Figure 3).

%% TODO: add figure

\textbf{Replication} is the process of duplicating DNA, which allows a cell to divide such that the resulting daughter cells retain complete copies of its genetic information. DNA consists of four different so-called nucleobases named adenine (A), cytosine (C), guanine (G), and thymine (T). Each strand of the double helix structure of DNA is a linear chain of nucleobases. The two strands are held together by hydrogen bonds since adenine can form three hydrogen bonds with thymine and cytosine can form two hydrogen bonds with guanine. Each nucleobase can only bond with one other nucleobase, making the strands complementary to one another. 
Since one strand can be deduced from its complementary strand, DNA can be represented by a sequence of letters, for example, "ACTCGGTTTAGCA".

A stretch of DNA that contains a genetic blueprint for a particular molecule is called a gene, and the collection of an organism's genes is called its genome. \textbf{Transcription} is the process of synthesising an RNA molecule from a gene, and the resulting molecule is called a transcript. RNA is similar to DNA but differs in several ways; most notably all thymine nucleobases are replaced with uracil (U), and RNA molecules consist of only one strand.
% HT: , which is able to exit the nucleus compartment.
% RC: I haven't introduced the concept of a nucleus yet
Due to its single-strandedness, RNA is less stable and will break down faster. Single-strandedness also allows some types of RNA (e.g. transfer RNA, ribosomal RNA) to form more complex three-dimensional structures by having certain regions bind to other regions of the strand. This work only considers messenger RNA (mRNA), which are transcribed from protein-coding genes, meaning that the mRNAs can result in the production of particular proteins. A protein-coding gene consists of alternating sections called introns and exons. Exonic regions contain code for what a protein should be made up of, whereas the intronic regions can contain information on how to assemble the pieces.
% TODO: mention that there are about 20'000 protein-coding genes in the human body.

\textbf{RNA splicing} is a process that occurs in almost all organisms and results in the removal of intronic regions in the mRNA molecules, resulting in a mature mRNA. 
Splicing allows to create multiple variants of the same product, which can affect the enzymatic properties or localisation of the resulting product \cite{kelemen_functionalternativesplicing_2013}.

During \textbf{translation}, a chain of amino acids is synthesised from a mature mRNA transcript. 
% HT: in ribosomes. These units can be seen as small factories in which amino acids are assembled following the pattern contained in an mRNA molecule.
% RC: I can avoid talking about ribosomes
Every three nucleobases are translated into one of 21 different amino acids. The resulting chain of amino acids is folded up into a protein, the structure of which is determined by the sequence of different amino acids in the chain. In turn, its structure determines the functionality of the protein, which includes catalysing biochemical reactions, providing structure, and transportation of molecules.

@@

\textit{Homo sapiens} like to categorise everything they encounter and so too have they conceptualised groups of cells called "cell types" according to their functionality. The concept of cell types eases reasoning about all aspects of biology, for instance, which cell types turn into (differentiate) or communicate with which other cell types, or how a cell type responds to a specific stimulation. Cells can be highly specialised toward performing a particular function (e.g. memory B cells accelerate immune response by remembering previously encountered pathogens), or they can maintain a strong ability to differentiate into other cell types.

@@

\subsection{Cell dynamics and gene regulation}
If cells are dynamic entities and can gradually produce the molecules needed to acquire new functionality, what is the process by which this happens? The mechanism by which this happens is called gene regulation. Some proteins (or other molecules such as micro RNAs) are capable of determining the rate at which a gene is transcribed (transcription rate). Such proteins are called transcription factors (TFs), and the genes they regulate are called their targets. Typically, one TF will regulate the transcription rate of many targets. % TODO: download database and calculate some statistics.

@@

The earliest approaches for measuring the abundance of particular molecules in \emph{single cells} used the preferred instrument of every stereotypical biologist: the microscope.
% HT: I wouldn't write that if it was my thesis, but then again, maybe it's only my personal choice
Since it was developed by Coons et al. in 1941, immunohistochemistry (IHC) has been instrumental in visualising antigen-antibody proteins \cite{coons_immunologicalpropertiesantibody_1941}. In many multicellular organisms, antibodies and antigens serve as crucial communication tools as part of the organism's immune system. A cell can present a particular type of antigen on its cell surface, which allows a particular type of antibody to bind to it.

@@

% HT: Is the section above needed? I felt like this 1.1.5 section was quite hard to read compared to the light previous sections.

@@

\begin{figure}[htb!]
	\centering
	A. \includegraphics[width=.45\linewidth]{fig/profiling_convergence} 
	% also figure with cell numbers
	C. \includegraphics[width=.45\linewidth]{fig/multimodal} 
	% TODO: change image
	\caption{
		A. \textbf{Convergence of "Omics" Biology and Single-Cell Biology.} Technology that allows researchers to obtain genome-wide information from single cells is extending the boundaries of a field that has thus far been limited to the analyses of a select gene in eukaryotes. Image from Junker and van Oudenaarden (2014) \cite{junker_everycellspecial_2014}. 
		B. 
		C. scmultiomics \cite{moudgil_multimodalscrnaseq_2019}. 
		% TODO: rewrite the description
	}
	\label{fig:profiling_convergence}
\end{figure}

@@

\begin{figure}[htb!]
	\centering\includegraphics[width=.5\linewidth]{fig/singlecell_tools}
	\centering\includegraphics[width=.3\linewidth]{fig/scrnatools}
	% TODO: change image
	\caption{\textbf{A. Single-cell omics allows for many new types of computational approaches.} Figure adapted from Wagner et al. (2016) \cite{wagner_revealingvectorscellular_2016}.
		B. Zappia et al. (2018) \cite{zappia_exploringsinglecellrnaseq_2018}
		% TODO: rewrite the description
	}
	\label{fig:comp_tools}
\end{figure}

@@

\begin{figure}[htb!]
	\centering\includegraphics[width=\Largefigure]{fig/trajectory_inference_slim} % TODO: Update figure to match text!!!!
	\caption{
		Applications of single-cell trajectory inference methods. (A) Single-cell omics data appropriate for TI can be both obtained from an unsynchronised population of single cells (snapshot data) but also from synchronised cell populations. (B) UPDATE! (C) UPDATE!
	}
	\label{fig:trajectory_inference}
\end{figure}

@@

\begin{figure}[htb!]
	\centering
	\includegraphics[width=\Largefigure]{fig/ni_bulk_vs_sc}
	\caption{Bulk expression data return the average expressions of genes among large numbers of cells. In order to infer regulatory networks from this type of data, multiple bulk profiles (resulting from time series or perturbation experiments) are required. On the other hand, sequencing the transcriptome at the single-cell level uncovers the high variability among cells, providing the necessary information to infer gene regulatory networks directly.} % TODO: update label
	\label{fig:bulk_vs_singlecell}
\end{figure}

@@

\subsection{Co-author publications}
\begin{itemize}
	\item Decock A, Ongenaert M, \textbf{Cannoodt R}, Verniers K, De Wilde B, Laureys G, Van Roy N, Berbegall AP, Bienertova-Vasku J, Bown N, Clément N. Methyl-CpG-binding domain sequencing reveals a prognostic methylation signature in neuroblastoma. Oncotarget. 2016 Jan 12;7(2):1960.
	\item Van Cauwenbergh C, Van Schil K, \textbf{Cannoodt R}, Bauwens M, Van Laethem T, De Jaegere S, Steyaert W, Sante T, Menten B, Leroy BP, Coppieters F. arrEYE: a customized platform for high-resolution copy number analysis of coding and noncoding regions of known and candidate retinal dystrophy genes and retinal noncoding RNAs. Genetics in Medicine. 2017 Apr;19(4):457.
	\item Claeys S, Denecker G, \textbf{Cannoodt R}, Kumps C, Durinck K, Speleman F, De Preter K. Early and late effects of pharmacological ALK inhibition on the neuroblastoma transcriptome. Oncotarget. 2017 Dec 5;8(63):106820.
	\item Depuydt P, Boeva V, Hocking TD, \textbf{Cannoodt R}, Ambros IM, Ambros PF, Asgharzadeh S, Attiyeh EF, Combaret V, Defferrari R, Fischer M. Genomic amplifications and distal 6q loss: novel markers for poor survival in high-risk neuroblastoma patients. JNCI: Journal of the National Cancer Institute. 2018 Mar 5;110(10):1084-93.
	\item Scott CL, T'Jonck W, \ldots, \textbf{Cannoodt R}, Saelens W \ldots, Guilliams M. The transcription factor ZEB2 is required to maintain the tissue-specific identities of macrophages. Immunity. 2018 Aug 21;49(2):312-25.
	\item Saelens W, \textbf{Cannoodt R}, Saeys Y. A comprehensive evaluation of module detection methods for gene expression data. Nature communications. 2018 Mar 15;9(1):1090.
	\item Todorov H, \textbf{Cannoodt R}, Saelens W, Saeys Y. Network Inference from Single-Cell Transcriptomic Data. In Gene Regulatory Networks 2019 (pp. 235-249). Humana Press, New York, NY..
	\item Van den Berge K, De Bezieux HR, Street K, Saelens W, \textbf{Cannoodt R}, Saeys Y, Dudoit S, Clement L. Trajectory-based differential expression analysis for single-cell sequencing data. BioRxiv. 2019 Jan 1:623397.
	\item Weber LM, Saelens W, \textbf{Cannoodt R}, Soneson C, Hapfelmeier A, Gardner PP, Boulesteix AL, Saeys Y, Robinson MD. Essential guidelines for computational method benchmarking. Genome biology. 2019 Dec;20(1):125.
\end{itemize}

@@

%library(dlstats)
%library(tidyverse)
%
%cran <- c("babelwhale", "diffusionMap", "dyndimred", "dynparam", "dynutils", "GillespieSSA", "GillespieSSA2", "incgraph", "princurve", "proxyC", "qsub", "SCORPIUS", "devtools", "mlr", "ParamHelpers", "remotes", "rlang")
%bioc <- c("ClusterSignificance", "monocle", "slingshot", "splatter")
%
%dl_cran <- dlstats::cran_stats(cran)
%dl_bioc <- dlstats::bioc_stats(bioc)
%
%dls <- bind_rows(dl_cran, dl_bioc %>% select(start, end, package, downloads = Nb_of_downloads)) %>% 
%mutate(role = ifelse(package %in% c("devtools", "mlr", "ParamHelpers", "remotes", "rlang", bioc), "Contributor", "Author"))
%
%min_date <- "2019-08-01"
%num_days <- as.integer(Sys.Date() - as.Date(min_date))
%summ <- dls %>% filter("2019-08-01" <= start) %>% group_by(package) %>% summarise(downloads = sum(downloads) / num_days * 365) %>% as.data.frame

\begin{table}[ht!]
	\caption{\textbf{Contibutions to open-source software.} Following abbreviations denote the role within the package: aut Author, ctb Contributor. Yearly download statistics are based on the number of downloads between 2019-08-01 and 2019-09-10. CRAN download statistics are retrieved from the Rstudio CRAN mirror only; other CRAN mirrors do not track download statistics. For Github repositories, no download statistics could be retrieved. } \label{tab:packages}

@@

		\cranpkg{babelwhale} & aut & CRAN & 3996 & Interacting with Docker and Singularity containers \\
		\cranpkg{diffusionMap} & aut & CRAN & 21361 & Implements diffusion map method of data parameterization, including creation and visualization of diffusion map \\

@@

		\cranpkg{dyndimred} & aut & CRAN & 5511 & Applying dimensionality reduction methods \\

@@

		\cranpkg{dynparam} & aut & CRAN & 3084 & Creating meta-information for parameters \\

@@

		\cranpkg{dynutils} & aut & CRAN & 5657 & Common functionality for the dynverse packages \\

@@

		\cranpkg{GillespieSSA} & aut & CRAN & 7546 & Gillespie's Stochastic Simulation Algorithm (SSA) \\
		\cranpkg{GillespieSSA2} & aut & CRAN & 6506 & Gillespie's Stochastic Simulation Algorithm for Impatient People \\

@@

		\cranpkg{incgraph} & aut & CRAN & 3175 & Incremental graphlet counting for network optimisation \\
		\githubpkg{rcannood}{lmds} & aut & Github & \notavailable & Landmark Multi-Dimensional Scaling \\
		\cranpkg{princurve} & aut & CRAN & 26'991 & Fits a principal curve in arbitrary dimension \\
		\cranpkg{proxyC} & aut & CRAN & 117'484 & Computes proximity in large sparse matrices \\
		\cranpkg{qsub} & aut & CRAN & 3193 & Running commands remotely on gridengine clusters \\
		\cranpkg{SCORPIUS} & aut & CRAN & 4772 & Inferring developmental chronologies from single-cell RNA sequencing data \\ \hline\hline
		\biocpkgl{ClusterSignificance}{Clus\-ter\-Sig\-nif\-i\-cance} & ctb & Bioc & 803 & Assess if class clusters in dimensionality reduced data representations have a separation different from permuted data \\
		\cranpkg{devtools} & ctb & CRAN & 3'775'350 & Tools to make developing R packages easier \\

@@

		\cranpkg{mlr} & ctb & CRAN & 142'605 & Machine Learning in R \\
		\biocpkg{monocle} & ctb & Bioc & 35'240 & Clustering, differential expression, and trajectory analysis for single-cell RNA-Seq \\
		\cranpkg{ParamHelpers} & ctb & CRAN & 109'408 & Helpers for Parameters in Black-Box Optimization, Tuning and Machine Learning \\

@@

		\cranpkg{remotes} & ctb & CRAN & 3'704'594 & R package installation from remote repositories, including GitHub \\
		\cranpkg{rlang} & ctb & CRAN & 11'470'763 & Functions for base types and core R and tidyverse features \\

@@

		\biocpkg{slingshot} & ctb & Bioc & 11'643 & Tools for ordering single-cell sequencing \\
		\biocpkg{splatter} & ctb & Bioc & 3741 & Simple simulation of single-cell RNA sequencing data \\
